Technologies of Semantic WEB as an environment of application development and integration

### 2022-11-27 paper-aicts-jcart.tex
Release corrected

This paper presents a view on the technologies as data representation for software development tools based on model analysis, and transformation, %and retrospection capabilities,
examples are provided.  %Three examples show the technologies usage, exposing our experience.
% \end{abstract}

@@

Briefly, SW technologies from a developer point of view include the following assets:
% Es sheist dass diest Enumrate ist Itemize????
\begin{enumerate}
\item Global identification of the entities via URI, IRI, URL (\{Universal, International\} Resource \{Identifier, Locator\}) are essentially the same notions but reflecting different representation properties of an identified resource;
\item Conceptual set-theoretic and category-theoretic data representation notation as relations between resources and literals represented % with triples of a form
as \texttt{<subject, predicate, object>};
\item Triples used in a description of a domain form a corresponding semantic graph, conditionally subdivided into terminological subgraph (T-Box) and instances %subgraph
(A-Box);
\item Graph presentation formats and techniques are implemented in files, databases, markups of documents, a data flow between applicants; \label{kg1}
\item There are %public semantic
databases comprising the generic domain objects descriptions, which are mostly A-Boxes (databases), for example, \texttt{DBPedia.org}, (formalized \texttt{Wikipedia.org}). % wikidata
\item For graphs being databases, query services are implemented to allow users to obtain its data by portions, as well as implement modifications;
\item API and libraries for processing files and documents with semantic data acquisition.% (\emph{e.g.} by extraction of the markup).
\item Software for graph data processing to improve its structure or induct %general relation
patterns;
\item Validation and verification services allow one to explore consistency and completeness of the conceptual domain models and their implementations in software. \label{kg2}
\end{enumerate}
Some of the assets % \ref{kg1}--\ref{kg2}
form knowledge graph (KG) technological aspect of the SW that gained a particular interest within R\&D in the last decade.

@@

Triples are a combination of two or three IRIs (\texttt{<subjectIRI, predicateIRI, object>}).  Subject and predicate are always IRIs, whereas objects are either IRI (a reference to a resource) or a literal.  Literals are values of basic types (\emph{e.g.}, number, string, date, time) usually accompanied with metadata (type designator or a language tag).  Types are defined in a vocabulary \texttt{<\url{http://www.w3.org/2001/XMLSchema}\#>}, having well-known namespace abbreviation \ns{xs}\footnote{Hereafter, all well-known namespace names will be written with a bold monospace font.}.  A language tag defines string values attribute, affiliating them to a cultural context.  It is recognized with KG tools and utilized at presentation layers of the software.

@@

Much R\&D efforts since 2001 \cite{tbl} were spent to develop SW data formats and their coupling with document representation.  The simplest way to provide a printed document with semantic representation is by pointing the user's software agent to an URL serving the semantic data.  However, this approach requires implementing two different procedures for document generation and its metadata serving.  That is why an RDFa\footnote{Abbreviation of Resource Description Framework attribute language} has been developed.  It allows developer of XHTML\footnote{HTML structures with strict XML restrictions obeyed} documents to enrich the document with its semantic layer.  The semantic markup is represented with special attributes and their values.  The hierarchical structure of XHTML is a matrix layer for the hierarchical structure of document entities' semantic relations.  XHTML/RDFa processing libraries extract the layer.  There are other formats having the same expression capabilities as RDF but used for other purposes: \texttt{N-triples}, \texttt{Turtle}, \texttt{Notation~3}, \texttt{JSON-LD}.  The last is an efficient way of coupling SW and various generic data storage, indexing, and processing services.

@@

A convenient approach of application construction is usage component architectures, where subsystems in runtime are components \emph{providing} interfaces using other components' service available via corresponding interfaces.  Zope component architecture\footnote{On-line book is at \url{https://muthukadan.net/docs/zca.html}}, Logtalk, React.js are used to \emph{implement} interfaces.  Python allows to wrap whole subsystems, \emph{e.g.}, Elastic search, in imperative paradigm, and Logtalk is chosen for the declarative counterpart.  Inter-process component communications is realized usually with \texttt{HTTP} protocol.%, representing data in XML and JSON formats.

Our R\&D resulted in a set of components and a general application architecture showed in Figure~\ref{fig:arch-services}.
\begin{figure}%[!b]
  \begin{center}
    \includegraphics[width=0.9\linewidth]{architecture-mda-lod-ext-general.pdf}
  \end{center}
  \begin{center}
      \textbf{Abbreviations}%\\[1ex]
    \end{center}
    \begin{quote}\scriptsize
      % T-Module is Transformation module,
      MDA is Model-Driven Architecture,
      CIM is Computationally Independent Model,
      PIM is Platform Independent Model,
      PSM is Platform Specific Model,
      T-Box is Terminological Box,
      KB is Knowledge Base,
      % A-Box is Instance Box,
      % NGS is Next-Generation Sequencing,
      DB is Database.
    \end{quote}
  \caption{Architecture of application instrumental environment as a set of components}
  \label{fig:arch-services}
\end{figure}
The kernel module of applications is KG server (B2) built on the base of ClioPatria \cite{cliopatria} with web-server and Pengines plug-ins.  ClioPatria is a web-application implemented in SWI-Prolog \cite{swi} being ISO-compliant Prolog implementation.  KG functions are implemented with SW package, supplied in standard SWI-Prolog distribution.  Pengines service enables SW web-applications to utilize logical inference on the server-side \cite{pengines} with browser JavaScript environment integration.  Pengine's knowledge base is extended with the programmer's Prolog modules.  Kernel module stores a common global part of A-~and T-boxes of applications, \emph{i.e.}, it is application database.  Other modules provide an application (C3). % with services.

@@

This project \cite{authoring} is aimed at the construction of an environment for document authoring out of user-editable template elements, database query results, LOD-compliant web pages, and other kind of a document content.  The target document is a web page where content is inferred while loaded by the browser.

@@

For document markup (document data representation), the following vocabularies are used:
\begin{itemize}
  \item Friend-of-a-friend (\textbf{foaf}), agent information: individuals, legal entities, program agents;
  \item Provenance (\textbf{prov}), references between documents, data origins, proofs, \emph{etc.};
  \item Dublin Core (\textbf{dc}), web page annotation mark up;
  \item DBPedia resource (\textbf{dbr}), references to instant objects and classes;
  \item Schema.org (\textbf{schema}), Google, Yandex, Yahoo, \emph{etc}. searchable objects, structural elements;
  \item The Bibliographic Ontology (\textbf{bibo}), literature reference mark up.
\end{itemize}

The general logical structure (sections, subsections, \emph{etc.}) are represented with Open Annotation (\textbf{oa}) vocabulary originally used for annotation of already existing content. The generated document conversion into PDF and other formats is carried out under the control of \textbf{oa}.  HTML5 default tag structures are used for visual sectioning and style definition. The \textbf{oa} vocabulary is also used to define parts of a document, which allow interactive editing after the document has been rendered.  The obtained changes are fixed in a commit as new resources, and the containing document obtains new IRI fixing all the set of resource changes.

@@

Project \cite{gisviewer} devoted to investigating capabilities of SW and KG technologies in representation spatially distributed data (GIS\footnote{Abbreviation of Geographical Information System}-data).  The fault data \cite{lunina,afs} are chosen as a subject for representation and publishing in this investigation.  In geology, researchers accumulate data obtained after event observations, \emph{e.g.}, earthquakes, landslides, by analyzing remote sensing data and results of field works.  The obtained data are processed and interpreted, setting new attributes to a fault or refinement of their values.  According to the geological research techniques, additional information is associated with attributes, clarifying their values.  Such clarification comprises precision characteristics, measurement conditions, reliability assessment, paper references, and published fault data.

@@

\bibitem{tbl} T.~Berners-Lee, J.~Hendler, O.~Lassila.  The semantic web: A new form of web content that is meaningful to computers will unleash a revolution of new possibilities.  Scientific American, May 2001.

\bibitem{hogan} A.~Hogan, E.~Blomqvist, M.~Cochez, C.~D’Amato \emph{et al}. Knowledge Graphs, 2020. URL:\url{https://arxiv.org/abs/2003.02320v5} (access date: 12-Dec-2021)

\bibitem{zont19} E.~Cherkashin, A.~Shigarov, V.~Paramonov. Representation of MDA transformation with logical objects.  Procs. of International Multi-Conference on Engineering, Computer and Information Sciences (SIBIRCON), Novosibirsk, Russia, 2019, pp.~0913--0918 \doi{10.1109/SIBIRCON48586.2019.8958008}

\bibitem{cliopatria} J.~Wielemaker, W.~Beek, M.~Hildebrand, J.~Ossenbruggen, ClioPatria: A SWI-Prolog infrastructure for the Semantic Web, Semantic Web, 2016, volume 7(5). pp.~529--541 \doi{10.3233/SW"=150191}

% \bibitem{swi} J.~Wielemaker, T.~Schrijvers, M.~Triska, T.~Lager. SWI-Prolog. Theory and Practice of Logic Programming, volume~12, No.~1-2, pp.~67--96, 2011, ISSN~1471-0684.

\bibitem{pengines} T.~Lager, J.~Wielemaker. Pengines: Web Logic Programming Made Easy.  Theory and Practice of Logic Programming, volume~14, No.~4-5, 2014, pp.~539--552. \doi{10.1017/S1471068414000192}

\bibitem{lod} C.~Bizer, T.~Heath, T.~Berners-Lee, Linked Data -- The Story So Far.  Int. J. Semantic Web Inf. Syst., volume~5, 2009, pp.~1--22. \doi{10.4018/jswis.2009081901}

\bibitem{aicts2020} E.~Cherkashin, A.~Shigarov. Construction techniques of Baikal microbiome research information-computational environment.  Procs. of 1st International Workshop on Advanced Information and Computation Technologies and Systems. Irkutsk, Russia, December 7-11, 2020. CEUR-WS.org, URL:\url{http://ceur-ws.org/Vol-2858/paper2.pdf} (access date: 12-Dec-2021)

\bibitem{authoring} E.~Cherkashin, A.~Shigarov, V.~Paramonov, A.~Mikhailov, Digital archives supporting document content inference, Procs.  of 42-nd International Convention on Information and Communication Technology Electronics and Microelectronics (MIPRO), May 20–24, 2019.  pp. 1037-1042. \doi{10.23919/MIPRO.2019.8757196}

\bibitem{gisviewer} E.~Cherkashin, O.~Lunina, L.~Demyanov, A.~Tsygankov. Web-GIS viewer for active faults data represented as a knowledge graph.  Procs. of 4th Scientific-practical Workshop Information Technologies: Algorithms, Models, Systems. Irkutsk, Russia, September 14, 2021. CEUR-WS.org, URL:\url{http://ceur-ws.org/Vol-2984/paper8.pdf} (access date: 12-Dec-2021)

% \bibitem{lunina} O.~V.~Lunina.  The digital map of the Pliocene–Quaternary crustal faults in the Southern East Siberia and the adjacent Northern Mongolia.  Geodynamics \& Tectonophysics.  volume~7(3), 2016.  pp.~407-434. (in Russian) \doi{10.5800/GT-2016-7-3-0215}
% \bibitem{afs} A.~A.~Gladkov, O.~V.~Lunina. Cartographic service ``Activetectonics''. URL:\url{http://activetectonics.ru/} (access date: 20-Sep-2021)

\bibitem{shigpaper} N.~O.~Dorodnykh, A.~Y.~Yurin, A.~O.~Shigarov. Conceptual Model Engineering for Industrial Safety Inspection Based on Spreadsheet Data Analysis.  In: Simian D., Stoica L. (eds) Modelling and Development of Intelligent Systems.  MDIS 2019.  Communications in Computer and Information Science, volume~1126, 2020, Springer, Cham. \doi{10.1007/978-3-030-39237-6_4}

%\bibitem{foss} M.~Leidig, R.~Teeuw. Free software: A review, in the context of disaster management.  International Journal of Applied Earth Observation and Geoinformation, 42 (2015), pp.~49-56. \doi{10.1016/j.jag.2015.05.012}.

%\bibitem{lgd} C.~Stadler, J.~Lehmann, K.~Hffner, S.~Auer. LinkedGeoData: A core for a web of spatial open data. Semantic Web 3 (2012) 333–354. \doi{10.3233/SW-2011-0052}

%\bibitem{geolink} M.~Cheatham, A.~Krisnadhi, R.~Amini, P.~Hitzler, \emph{et al}. The GeoLink knowledge graph, Big Earth Data, 2:2 (2018), pp.~131-143. \doi{10.1080/20964471.2018.1469291}

%\bibitem{abid} T.~Abid, H.~Zarzour. Integrating linked open data in a geographical information system.  Procs.  of. International Conference on Information Technology for Organization Development. Oct 19-20, 2014, University of Tebessa, Tebessa, Algeria (2014).

%\bibitem{iwaniak1} A.~Iwaniak, I.~Kaczmarek, M.~Strzelecki, J.~Lukowicz, P.~Jankowski. Enriching and improving the quality of linked data with GIS. Open Geosciences, Vol.~8, 1~(2016) pp.~323-336. \doi{10.1515/geo-2016-0020}

%\bibitem{iwaniak17} A.~Iwaniak, M.~Leszczuk, M.~Strzelecki, F.~Harvey, I.~Kaczmarek. A Novel Approach for Publishing Linked Open Geodata from National Registries with the Use of Semantically Annotated Context Dependent Web Pages.  International Journal of Geo-Information. 6, 252 (2017). \doi{10.3390/ijgi6080252}

\end{thebibliography}
% \section{Online Resources}

% The sources for the viewer are being developed at Github, URL:~\url{https://github.com/De17eon/GRL}.

\end{document}

>>>>

This paper presents a view on the technologies as data representation for software development tools based on model analysis, and transformation, %and retrospection capabilities,
examples are provided.  %Three examples show the technologies' usage, exposing our experience.
% \end{abstract}

@@

Briefly, SW technologies from a developer point of view include the following assets:
% Es sheist dass diest Enumrate ist Itemize????
\begin{enumerate}
\item Global identification of the entities via URI, IRI, URL (\{Universal, International\} Resource \{Identifier, Locator\}) are essentially the same notions but reflecting different representation properties of an identified resource;
\item Conceptual set-theoretic and category-theoretic data representation notation as relations between resources and literals represented % with triples of a form
as \texttt{<subject, predicate, object>};
\item Triples used in a description of a domain form a corresponding semantic graph, conditionally subdivided into terminological subgraph (T-Box) and instances %subgraph
(A-Box);
\item Graph presentation formats and techniques are implemented in files, databases, markups of documents, a data flow between applicants; \label{kg1}
\item There are %public semantic
databases comprising the generic domain objects descriptions, which are mostly A-Boxes (databases), for example, \texttt{DBPedia.org}, (formalized \texttt{Wikipedia.org}). % wikidata
\item For graphs being databases, query services are implemented to allow users to obtain its data by portions, as well as implement modifications;
\item API and libraries for processing files and documents with semantic data acquisition.% (\emph{e.g.} by extraction of the markup).
\item Software for graph data processing to improve its structure or induct %general relation
patterns;
\item Validation and verification services allow one to explore consistency and completeness of the conceptual domain models and their implementations in software. \label{kg2}
\end{enumerate}
Some assets % \ref{kg1}--\ref{kg2}
form knowledge graph (KG) technological aspect of the SW that gained a particular interest within R\&D in the last decade.

@@

Triples are a combination of two or three IRIs (\texttt{<subjectIRI, predicateIRI, object>}).  Subject and predicate are always IRIs, whereas objects are either IRI (a reference to a resource) or a literal.  Literals are values of basic types (\emph{e.g.}, number, string, date, time) usually accompanied by metadata (type designator or a language tag).  Types are defined in a vocabulary \texttt{<\url{http://www.w3.org/2001/XMLSchema}\#>}, having well-known namespace abbreviation \ns{xs}\footnote{Hereafter, all well-known namespace names will be written with a bold monospace font.}.  A language tag defines string values attribute, affiliating them to a cultural context.  It is recognized with KG tools and utilized at presentation layers of the software.

@@

Much R\&D efforts since 2001 \cite{tbl} were spent to develop SW data formats and their coupling with document representation.  The simplest way to provide a printed document with semantic representation is by pointing the user's software agent to a URL serving the semantic data.  However, this approach requires implementing two different procedures for document generation and its metadata serving.  That is why an RDFa\footnote{Abbreviation of Resource Description Framework attribute language} has been developed.  It allows developer of XHTML\footnote{HTML structures with strict XML restrictions obeyed} documents to enrich the document with its semantic layer.  The semantic markup is represented with special attributes and their values.  The hierarchical structure of XHTML is a matrix layer for the hierarchical structure of document entities' semantic relations.  XHTML/RDFa processing libraries extract the layer.  There are other formats having the same expression capabilities as RDF but used for other purposes: \texttt{N-triples}, \texttt{Turtle}, \texttt{Notation~3}, \texttt{JSON-LD}.  The last is an efficient way of coupling SW and various generic data storage, indexing, and processing services.

@@

A convenient approach of application construction is usage component architectures, where subsystems in runtime are components \emph{providing} interfaces using other components' service available via corresponding interfaces.  Zope component architecture\footnote{On-line book is at \url{https://muthukadan.net/docs/zca.html}}, Logtalk, React.js are used to \emph{implement} interfaces.  Python allows wrapping whole subsystems, \emph{e.g.}, Elastic search, in imperative paradigm, and Logtalk is chosen for the declarative counterpart.  Inter-process component communications is realized usually with \texttt{HTTP} protocol.%, representing data in XML and JSON formats.

Our R\&D resulted in a set of components and a general application architecture showed in Figure~\ref{fig:arch-services}.
\begin{figure}%[!b]
  \begin{center}
    \includegraphics[width=0.9\linewidth]{architecture-mda-lod-ext-general.pdf}
  \end{center}
  \begin{center}
      \textbf{Abbreviations}%\\[1ex]
    \end{center}
    \begin{quote}\scriptsize
      % T-Module is Transformation module,
      MDA is Model-Driven Architecture,
      CIM is Computationally Independent Model,
      PIM is Platform Independent Model,
      PSM is Platform Specific Model,
      T-Box is Terminological Box,
      KB is Knowledge Base,
      % A-Box is Instance Box,
      % NGS is Next-Generation Sequencing,
      DB is Database.
    \end{quote}
  \caption{Architecture of application instrumental environment as a set of components}
  \label{fig:arch-services}
\end{figure}
The kernel module of applications is KG server (B2) built on the base of ClioPatria \cite{cliopatria} with web-server and Pengines plug-ins.  ClioPatria is a web-application implemented in SWI-Prolog being ISO-compliant Prolog implementation.  KG functions are implemented with SW package, supplied in standard SWI-Prolog distribution.  Pengines service enables SW web-applications to utilize logical inference on the server-side \cite{pengines} with browser JavaScript environment integration.  Pengine's knowledge base is extended with the programmer's Prolog modules.  Kernel module stores a common global part of A-~and T-boxes of applications, \emph{i.e.}, it is application database.  Other modules provide an application (C3). % with services.

@@

This project \cite{authoring} is aimed at the construction of an environment for document authoring out of user-editable template elements, database query results, LOD-compliant web pages, and other document content.  The target document is a web page where content is inferred while loaded by the browser.

@@

For document markup (document data representation), the following vocabularies are used:
\begin{itemize}
  \item Friend-of-a-friend (\textbf{foaf}), agent information: individuals, legal entities, program agents;
  \item Provenance (\textbf{prov}), references between documents, data origins, proofs, \emph{etc.};
  \item Dublin Core (\textbf{dc}), web page annotation mark up;
  \item DBPedia resource (\textbf{dbr}), references to instant objects and classes;
  \item Schema.org (\textbf{schema}), Google, Yandex, Yahoo, \emph{i.e.}, searchable objects, structural elements;
  \item The Bibliographic Ontology (\textbf{bibo}), literature reference mark up.
\end{itemize}

The general logical structure (sections, subsections) is represented with Open Annotation (\textbf{oa}) vocabulary originally used for annotation of already existing content. The generated document conversion into PDF and other formats is carried out under the control of \textbf{oa}.  HTML5 default tag structures are used for visual sectioning and style definition. The \textbf{oa} vocabulary is also used to define parts of a document, which allow interactive editing after the document has been rendered.  The obtained changes are fixed in a commit as new resources, and the containing document obtains new IRI fixing all the set of resource changes.

@@

Project \cite{gisviewer} devoted to investigating capabilities of SW and KG technologies in representation spatially distributed data (GIS\footnote{Abbreviation of Geographical Information System}-data).  The fault data \cite{gisviewer} are chosen as a subject for representation and publishing in this investigation.  In geology, researchers accumulate data obtained after event observations, \emph{e.g.}, earthquakes, landslides, by analyzing remote sensing data and results of field works.  The obtained data are processed and interpreted, setting new attributes to a fault or refinement of their values.  According to the geological research techniques, additional information is associated with attributes, clarifying their values.  Such clarification comprises precision characteristics, measurement conditions, reliability assessment, paper references, and published fault data.

@@

\bibitem{tbl} Berners-Lee~T., Hendler~J., Lassila~O.  The semantic web: A new form of web content that is meaningful to computers will unleash a revolution of new possibilities.  Scientific American. 2001; 284:~1--5

\bibitem{hogan} Hogan~A., Blomqvist~E., Cochez~M., D’Amato~C., \emph{et al}. Knowledge Graphs. 2020. Available~at:~\url{https://arxiv.org/abs/2003.02320v5} (Accessed December 12, 2021)

\bibitem{zont19} Cherkashin~E., Shigarov~A., Paramonov~V. Representation of MDA transformation with logical objects.  Proc. of ``International Multi-Conference on Engineering, Computer and Information Sciences (SIBIRCON)''. Novosibirsk, Russia; 2019:~0913--0918 \doi{10.1109/SIBIRCON48586.2019.8958008}

\bibitem{cliopatria} Wielemaker~J., Beek~W., Hildebrand~M., Ossenbruggen~J. ClioPatria: A SWI-Prolog infrastructure for the Semantic Web. Semantic Web. 2016; 7(5):~529--541 \doi{10.3233/SW-150191}

\bibitem{pengines} Lager~T., Wielemaker~J. Pengines: Web Logic Programming Made Easy.  Theory and Practice of Logic Programming. 2014; 14(4-5):~539--552. \doi{10.1017/S1471068414000192}

\bibitem{lod} Bizer~C., Heath~T., Berners-Lee~T., Linked Data -- The Story So Far.  Int. J. Semantic Web Inf. Syst. 2009; 5:~1--22. \doi{10.4018/jswis.2009081901}

\bibitem{aicts2020} Cherkashin~E., Shigarov~A. Construction techniques of Baikal microbiome research information-computational environment.  Proc. of ``1st International Workshop on Advanced Information and Computation Technologies and Systems''. December 7--11. Irkutsk, Russia. CEUR-WS.org; 2020. Available~at:~\url{http://ceur-ws.org/Vol-2858/paper2.pdf} (Accessed December 12, 2021)

\bibitem{authoring} Cherkashin~E., Shigarov~A., Paramonov~V., Mikhailov~A., Digital archives supporting document content inference. Proc. of ``42-nd International Convention on Information and Communication Technology Electronics and Microelectronics (MIPRO)''. May 20--24. Opatia, Chroatia; 2019; 1037--1042. \doi{10.23919/MIPRO.2019.8757196}

\bibitem{gisviewer} Cherkashin~E., Lunina~J., Demyanov~L., Tsygankov~A. Web--GIS viewer for active faults data represented as a knowledge graph.  Proc. of ``4th Scientific-practical Workshop Information Technologies: Algorithms, Models, Systems''. September 14. Irkutsk, Russia, CEUR-WS.org; 2021. Available~at:~\url{http://ceur-ws.org/Vol-2984/paper8.pdf} (Accessed December 12, 2021)

\bibitem{shigpaper} Dorodnykh~N., Yurin~A., Shigarov~A. Conceptual Model Engineering for Industrial Safety Inspection Based on Spreadsheet Data Analysis.  In: Simian D., Stoica L. (eds) Modelling and Development of Intelligent Systems.  MDIS 2019.  Communications in Computer and Information Science. Springer, Cham. 2020; 1126.  \doi{10.1007/978-3-030-39237-6_4}

\end{thebibliography}
% \section{On-line Resources}

% The sources for the viewer are being developed at GitHub, URL:~\url{https://github.com/De17eon/GRL}.
\label{pageend}
\end{document}
